---
title: "Lubridate: Theory and Practice"
author: "Laura Noetzel"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# Overview  

- Lubridate is a R package that belongs to the tidyverse and makes working with dates and times easier and more intuitive  
- Base R functions for dates and times are unintuitive and limited such as `as.POSIXct()`, `ISOdate()` or `as.POSIXct.Date()`  
- Lubridate  
  + offers consistent and easy to handle functions, that take special features as leap years, leap seconds, daylight savings time (DST) or time zones into account  
  + overrides base R functions for working with dates and times  
  + focuses on working with dates and datetimes, not times in particular (for that see hms package)  
  + refers to a specific moment in time as an **instant**  
- A [Cheat Sheet](https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf) with the most important formulas and explanations is available  

# Prerequisites  

- Lubridate is not part of the core tidyverse and thus needs an extra installation:  
```{r}
install.packages("lubridate")
```
- When working with lubridate installing and using the tidyverse is recommended:  
```{r}
library(lubridate)
library(tidyverse)
```
- The Development version of the package is available on GitHub (other packages needed to install)  

## Preparing the Data

- 
