---
title: "Lubridate: Theory and Practice"
author: "Laura Noetzel"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="99999")
opts_chunk$set(echo=T,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=99999)
```

# Overview  

- Lubridate is a R package that belongs to the tidyverse and makes working with dates and times easier and more intuitive  

- Base R functions for dates and times are unintuitive and limited such as `as.POSIXct()`, `ISOdate()` or `as.POSIXct.Date()`  

- Lubridate  
  + Offers consistent and easy to handle functions, that take special features as leap years, leap seconds, daylight savings time (DST) or time zones into account  
  + Overrides base R functions for working with dates and times  
  + Focuses on working with dates and datetimes, not times in particular (for that see hms package)  
  + Refers to a specific moment in time as an **instant**      


- A [Cheat Sheet](https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf) with the most important formulas and explanations is available  

# Prerequisites  

- Lubridate is not part of the core tidyverse and thus needs an extra installation via `install.packages("lubridate")`  

- When working with lubridate installing and using the tidyverse is recommended:  

```{r}
library(lubridate)
library(tidyverse)
```  

- The Development version of the package is available on GitHub (other packages needed to install)  

## Preparing the Data  

- Load and inspect the data, check for NAs and remove empty columns  

```{r}
df <- read_csv("police_project.csv")
View(df)

df %>%
  count(is.na(county_name))
df %>%
  count(is.na(search_type))

df <- df %>%
  within(rm(county_name))
```  

- Set theme for visualizations to classic  

```{r}
theme_set(theme_classic())
```

# Creating Dates and Datetimes  

- Three types refer to an instant  
  + Date `<date>`   
    - If displayed in numeric context (e.g. histogram) 1 means one day  
  + Time `<time>`   
  + Datetime `<dttm>`  
    - If displayed in numeric context (e.g. histogram) 1 means one second  

- Always use *simplest* possible type (no unnecessary datetime)    

- Get current date or datetime via `today()` and `now()` functions  

## Create Dates and Datetimes from String  

- Functions automatically work out format if order of **y**ear, **m**onth, **d**ay, **h**our, **m**inute and **s**econd 
  + Automatically recognize common separators **"-"**, **"/"**, **"."** and **" "**  
  
```{r}
mdy_hm("January, 31st, 2019 10:45")
```  

## Create Dates and Datetimes from Individual Components/Columns  

- For making dates use `make_date()`  

- For making datetimes use `make_datetime()`  

```{r}
# New column with format datetime
df <- df %>%
  mutate(stop_date_time=make_datetime(year=year(stop_date), month = month(stop_date), day = 
         day(stop_date), hour = hour(stop_time), min = minute(stop_time), sec = second(stop_time)))
```

- In this case very long, instead the functions from the chapter before can be used  

```{r}
df <- df %>%
  mutate(stop_date_time=ymd_hms(paste(stop_date,stop_time)))
```

## Create Dates and Datetimes from Other Types  

- For switching between date and datetime or when working with numeric offsets from unix systems use `as_date()` or `as_datetime()`  

# Components of Dates and Datetimes  

## Getting Components  

- Use accessor functions to get specific components of dates or datetimes, similar to functions creating dates and datetimes, such as `year()`, `month()`, `yday()`, `mday()`, `wday()`, `hour()`, `minute()`, `second()`  
  + For `month()` and `wday()` set **label = T** to display the name and set **abbr = F** to display the full name, otherwise the abbrevation will be shown  
  
```{r}
# Show all speeding drivers under 25 while grouping by year component of stop_date_time
df %>% 
  group_by(year(stop_date_time)) %>% 
  filter(violation=="Speeding" & driver_age < 25) %>% 
  count()
```
```{r}
# Show the overall number of drunk drivers sorted by weekday component of stop_date_time
df %>% 
  filter(drugs_related_stop == T) %>%
  group_by(weekday=wday(stop_date_time, label = T, abbr = F, week_start = 1)) %>%
  count() %>%
  ggplot(aes(x = weekday, y = n)) +
  geom_bar(stat = "identity", aes(fill = n)) +
  scale_fill_gradient(low = "palevioletred", high = "palevioletred4")
```  

## Setting Components  

- To set a component of a single date or datetime to a specific value use the same functions as for getting components, e.g. `year()`, `month()`, `hour()` etc.  

- To set multiple values at once use `update()`, where all date components for a certain variable or column can be specified at once  

- If a set value is too big, it will automatically roll over to the next bigger unit, e.g. adding 30 days to Feb. 1st will result in a date or datetime with month march  

## Rounding Components  

- Rounding dates is useful when dealing with units that are not a direct component of a date or datetime, such as quarters or weeks  

- Three functions to round  
  + `floor_date()` rounds to the next smaller unit  
  + `round_date()` rounds to mathematically nearest unit  
  + `ceiling_date()` rounds to next bigger unit  
  
```{r}
# New columns for rounded quarters
df <- df %>% 
    mutate(Quarter = as_date(round_date(x = stop_date_time, unit = "quarter")),
           Floor_Quarter = as_date(floor_date(stop_date_time, "quarter")),
           Ceiling_Quarter = as_date(ceiling_date(stop_date_time, "quarter")))
df[1,] %>% select(stop_date_time, Quarter, Floor_Quarter, Ceiling_Quarter)
```  
```{r}
# Show the number of drug related stops in 2015 per week
df %>%
  filter(year(stop_date_time)==2015, drugs_related_stop == T) %>%
  count(week = floor_date(stop_date_time, "week", week_start = 1)) %>%
  ggplot(aes(x = week, y = n)) +
  geom_line()
```  

# Time Spans  

- Three classes of time spans in R  

## Duration  

- Durations represent the exact **number of seconds** between two instants (fixed length)  

- All difftime objects can be formatted into a duration to get a fixed result unit (seconds) with `as.duration()`  

```{r}
# Calculating the time difference between two stops
df <- df %>% 
  mutate(time_diff =stop_date_time - lag(stop_date_time),
         duration = as.duration(time_diff))
df %>% select(stop_date_time, duration)
```  

- Calculating with durations works with similar functions as before, e.g. `dyears()` or `dhours()`  

```{r}
# Calculate limitation date
df <- df %>% 
  mutate(limitation = as_date(stop_date_time + dyears(3)))
df %>% select(stop_date_time, limitation)
```  

**ATTENTION**  

- No `dmonth()` function as months don't have the same length  

- When confronted with higher units than seconds the conversion rate is 60 seconds a minute, 60 minutes an hour etc. and thus specalities such as DST aren't considered  

## Period  

- Periods don't have the same length, but work in human units such as days, months etc.  

- They are more intuitive to work with than durations  

- Takes specialities such as DST or leap years into account  

- All difftime objects can be formatted into a period with `as.period()`  

```{r}
# Calculating the time difference between two stops as a period
df <- df %>% 
  mutate(period = as.period(time_diff))
df %>% select(stop_date_time, duration, period)
```  

- Creating and calculating with periods works with similar functions as before, e.g. `years()`, `months()`, `hours()` etc.  

```{r}
# Calculate limitation date with period
df <- df %>% 
  mutate(limitation_period = as_date(stop_date_time + years(3)))
df %>% select(stop_date_time, limitation, limitation_period)
```  

**ATTENTION**  

- `hour()` != `hours()`, `month()` != `months()` etc.  

## Interval  

- Intervals represent a duration with a specific starting point (and ending point)  

- Accurate measurement of time spans (periods and durations lack context often)  

- Not useful for datetime calculations as intervals are bound to a start and an end  
  + If intervals are added to datetime objects R coerces it into a duration  

- When building an interval the date that occurs earlier in time must be called first, so that the length is always positive  

```{r}
df <- df %>% mutate(interval = lag(stop_date_time) %--% stop_date_time)
df %>% select(stop_date_time, interval)

# Or use

df <- df %>% 
  mutate(interval = interval(start = lag(stop_date_time), end = stop_date_time))
```  

- Calculate the periods in an interval with integer division `%/%` (division without remainers)  

```{r}
df <- df %>% 
  mutate(interval_days = interval %/% days(1))
df %>% select(stop_date_time, interval_days)
```  

## What Time Span to Use  

- Always use simplest possible data structure  
  + Use **duration** if you only care for physical time  
  + Use **period** if you need to work with human units  
  + Use **interval** if you need to know how long a time span is in human units  

# Time Zones  

- Time zones are complicated to work with, because of geopolitical entities or ambiguous everyday names, such as Eastern Standard Time, which is different in the USA, Canada or Australia  

- Time Zone is an attribute of the datetime object  

- R uses international standard time zones displayed in `<continent>/<city>`  

- If no time zone is specified R assumes UTC, which doesn't use DST  

- Operations that combine dates or datetimes (e.g. `c()`) often drop assigned time zone and use the local time zone  

- Local time zone can be displayed via `Sys.timezone()`  

- Complete list of time zones can be displayed via `OlsonNames()`  

- Change time zone on two levels  
  + Keep the instant and only change how it's displayed `with_tz()` 
  + Change the underlying instant, e.g. when wrong time zone is assigned `force_tz()`  

```{r}
# Keep instant
df <- df %>% 
  mutate(displayed_tz = with_tz(time = stop_date_time, tzone = "America/New_York"))
# Change instant
df <- df %>% 
  mutate(changed_tz = force_tz(time = stop_date_time, tzone = "America/New_York"))
df %>% select(stop_date_time, displayed_tz, changed_tz)
```  

# Sources  

1. https://www.kaggle.com/faressayah/stanford-open-policing-project  
2. https://openpolicing.stanford.edu/  
3. https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf  
4. https://r4ds.had.co.nz/dates-and-times.html  
5. https://www.rdocumentation.org/packages/lubridate/versions/1.7.4/topics/interval  
6. https://lubridate.tidyverse.org/reference/round_date.html  
7. https://lubridate.tidyverse.org/  
8. https://www.jstatsoft.org/article/view/v040i03  
